---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 6 - Due date 03/06/23"
author: "Taro Katayama"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

## Questions 

This assignment has general questions about ARIMA Models. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(tidyverse)
#install.packages("sarima")
library(sarima)
```


## Q1
Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models: 


* AR(2)

> Answer: ACF: Will decay exponentially with time.
          PACF: Will identify the order of the AR model. The model will cut off at lag 2. 

* MA(1)

> Answer: ACF: Will identify the order of the model. The model will cut off at lag 1.
          PACF: Will decay exponentially with time.

## Q2
Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models

```{r}
#ARMA(1,0)
ARMAmodel_1<- arima.sim(model=list(ar=0.6), n=100) #the AR coefficient is 0.6

#ARMA(0,1)
ARMAmodel_2<- arima.sim(model=list(ma=0.9), n=100) #the MA coefficient is 0.9

#ARMA(1,1)
ARMAmodel_3<- arima.sim(model=list(ar=0.6,ma=0.9), n=100)
```


(a) Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r}
par(mfrow=c(1,3))
Acf(ARMAmodel_1,lag=40, ylab="ACF")
Acf(ARMAmodel_2,lag=40)
Acf(ARMAmodel_3,lag=40)
```


(b) Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r}
par(mfrow=c(1,3))
Pacf(ARMAmodel_1,lag=40)
Pacf(ARMAmodel_2,lag=40)
Pacf(ARMAmodel_3,lag=40)
```

(c) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer:
For ARMAmodel1, I would predict that it is an AR model with an order of 1, since the ACF has a slow decay and the PACF cuts off after lag 1. 
For ARMAmodel2, I would predict it is an ARMA model, with an order of 1 for the MA component. This is because the ACF cuts off after lag 1. When looking at the PACF, it has a geometric decaying pattern, which leads me to suspect it is an ARMA model.
For ARMAmodel3, I would predict it is an ARMA model. The ACF component looks to decay, which potentially suggests an AR model, but when looking at the PACF, the shape is geometric. The geometric shape of the PACF leads me to believe that it is an ARMA model.

(d) Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: They do not match for all three models. The coefficient for the AR model does not match the value of the PACF. For the second model, you cannot tell the MA coefficient. The theoretical value for MA is not displayed in ACF or PACF. The third model also does not match, and is not possible to tell the MA coefficient.

(e) Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r}
#ARMA(1,0)
ARMAmodel_1.1<- arima.sim(model=list(ar=0.6), n=1000) #the AR coefficient is 0.6

#ARMA(0,1)
ARMAmodel_2.1<- arima.sim(model=list(ma=0.9), n=1000) #the MA coefficient is 0.9

#ARMA(1,1)
ARMAmodel_3.1<- arima.sim(model=list(ar=0.6,ma=0.9), n=1000)
```
Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r}
par(mfrow=c(1,3))
Acf(ARMAmodel_1.1,lag=40)
Acf(ARMAmodel_2.1,lag=40)
Acf(ARMAmodel_3.1,lag=40)
```


(b) Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r}
par(mfrow=c(1,3))
Pacf(ARMAmodel_1.1,lag=40)
Pacf(ARMAmodel_2.1,lag=40)
Pacf(ARMAmodel_3.1,lag=40)
```

(c) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer:
For ARMAmodel1.1, I would predict that it is an AR model with an order of 1, since the ACF has a slow decay and the PACF cuts off after lag 1. 
For ARMAmodel2, I would predict it is an ARMA model, with an order of 1 for the MA component. This is because the ACF cuts off after lag 1. When looking at the PACF, it has a geometric decaying pattern, which leads me to suspect it is an ARMA model.
For ARMAmodel3, I would predict it is an ARMA model. The ACF component looks to decay, which potentially suggests an AR model, but when looking at the PACF, the shape is geometric. The geometric shape of the PACF leads me to believe that it is an ARMA model.


(d) Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: They do not match for all three models. The coefficient for the AR model does not match the value of the PACF. However, the value of the PACF is much closer than before to the coefficient, which suggests that with an increase in the number of observations, the value of the PACF gets closer to the coefficient. For the second model, you cannot tell the MA coefficient. The theoretical value for MA is not displayed in ACF or PACF. The third model does not match the coefficient for the AR model to the value of the PACF, nor can you derive the coefficient value of MA from PACF or ACF. 


## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$


(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.
>ARIMA(1,0,1)(1,0,0)_12
p=1 because y_{t-1} is added
d=0 because there is no constant value
q=1 because a_{t-1} is added
P=1 because y_{t-12} is added
D=0 because there is no constant value
s=12 because y_{t-12} shows that s=12

(b) Also from the equation what are the values of the parameters, i.e., model coefficients. 
>phi1=0.7 (precedes the y_{t-1} term), theta=0.1 (precedes the a_{t-1} term), phi12=-0.25 (precedes the y_{t-12} term)

## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}
SARIMAmodel_4<- sim_sarima(model=list(ma=0.5,sar=0.8, nseasons=12), n=1000)
par(mfrow=c(1,2))
Acf(SARIMAmodel_4,lag=40)
Pacf(SARIMAmodel_4,lag=40)
```
>You can tell the order of the seasonal component based on the spikes of ACF every 12 lags, which means P=1. I was not able to identify the order of the non-seasonal component. The ACF and PACF have unusual patterns which makes it hard to tell what the non-seasonal order is. 