---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 2 - Due date 02/03/23"
author: "Taro Katayama"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
#install.packages("tseries")
#install.packages("dplyr")
library(dplyr)
library(tseries)
library(lubridate)
library(ggplot2)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2022 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r}
#Importing data set
table_10.1<- read.table(file= "./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv",
                        header=TRUE,dec = ".",sep=",",stringsAsFactors = TRUE)
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
table_10.1<- table_10.1[, c("Month", "Total.Biomass.Energy.Production", "Total.Renewable.Energy.Production", "Hydroelectric.Power.Consumption")]
head(table_10.1)
str(table_10.1)
date<- ym(table_10.1$Month)
format(date,format="%Y-%m")
table_10.1<-cbind(date, table_10.1[,2:4])
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
?ts()
ts_table_10.1<- ts(table_10.1[,2:4], start = 1, end = 597, frequency = 1)
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
mean(ts_table_10.1[,1])
sd(ts_table_10.1[,1])
mean(ts_table_10.1[,2])
sd(ts_table_10.1[,2])
mean(ts_table_10.1[,3])
sd(ts_table_10.1[,3])
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
ggplot(table_10.1, aes(x=date, y=Total.Biomass.Energy.Production)) +
            geom_line(color="blue")+ geom_line(aes(y= mean(Total.Biomass.Energy.Production)), color = "red")+
  xlab("Date")+ ylab("Total Biomass Energy Production")+
  theme_bw()
ggplot(table_10.1, aes(x=date, y=Total.Renewable.Energy.Production)) +
            geom_line(color="blue")+ geom_line(aes(y= mean(Total.Renewable.Energy.Production)), color = "purple")+
  xlab("Date")+ ylab("Total Renewable Energy Production")+
  theme_bw()
ggplot(table_10.1, aes(x=date, y=Hydroelectric.Power.Consumption)) +
            geom_line(color="blue")+ geom_line(aes(y= mean(Hydroelectric.Power.Consumption)), color = "green")+
  xlab("Date")+ ylab("Hydroelectric Power Consumption")+
  theme_bw()
```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
?cor.test
cor.test(ts_table_10.1[,1], ts_table_10.1[,2])
#total biomass production and total renewable energy production are significantly correlated. Since biomass is a renewable energy source, the rise in biomass production also relates to total renewable energy production. However, just because these two variables are correlated, it does not mean that the rise in total renewable energy is caused by a rise in biomass production.

cor.test(ts_table_10.1[,2], ts_table_10.1[,3])
#there is a slight negative correlation between total renewable energy production and hydroelectric power consumption of -0.100. This correlation is significant with a p value of <0.05. 
cor.test(ts_table_10.1[,1], ts_table_10.1[,3])
#There is a moderate negative correlation between total biomass production and hydroelectric power consumption of -0.300. This correlation is significant with a p value of <0.05. 
```


## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
acf(ts_table_10.1[,1],lag.max = 40)
acf(ts_table_10.1[,2],lag.max = 40)
acf(ts_table_10.1[,3], lag.max = 40)
#total biomass production and total renewable energy production have similar behaviors. Since both of these variables increased with time, as you go further out in the lag, the autocorrelation decreased. 
#hydroelectric consumption had a cyclical autocorrelation as the lags increased. This means that hydroelectric consumption has a seasonality to it. 
```


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
pacf(ts_table_10.1[,1],lag.max = 40)
pacf(ts_table_10.1[,2],lag.max = 40)
pacf(ts_table_10.1[,3], lag.max = 40)

#these plots differ in that there is less of a noticeable trend or pattern with the pacf compared to acf.
```


